---
title: "PSMA PET SUVmax and its correlates: reproducible analysis"
subtitle: "Motchoffo Simo et al., The Prostate"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 7.2, dpi = 150)
options(width = 110)
```

# Purpose

This file reproduces **every number** reported in the manuscript and the supplement from a
single input, `PSMA_PET_source_data.xlsx`. Nothing is hard-coded: each value quoted in the
paper is recomputed here, and the final section asserts the published values so that a failed
`stopifnot` is the signal that text and data have drifted apart.

| Section | Reproduces |
|---|---|
| 2 | Data load and derived variables |
| 3 | Table 1 and Results "Cohort" |
| 4 | Table 2 and Results "Uptake tracks tumor architecture"; Figures 1 and 2 |
| 5 | Detectable effect size; adjusted models; Figure 3 |
| 6 | Digital pathology; Figure 4 and Figure S3 |
| 7 | Supplemental Results: secondary PET metrics (Table S2) |
| 8 | Supplemental Results: foci below the SUVmax threshold (Table S4, Figure S4) |
| 9 | Supplemental Results: anatomic location and multifocality |
| 10 | Session information and assertions |

**Statistical conventions.** The analysis is exploratory. Spearman rank correlation is used for
continuous covariates and the Kruskal-Wallis or Mann-Whitney test for categorical ones, all at
base-R defaults. **No correction for multiple comparisons is applied anywhere**, deliberately:
nothing here is presented as confirmatory, and p values are reported to convey the strength of
individual associations.

```{r packages}
library(readxl); library(dplyr); library(irr)
SOURCE_FILE    <- "PSMA_PET_source_data.xlsx"
PET_LOW_CUTOFF <- 4     # pre-specified in the parent-trial protocol
```

# Data load and derived variables

```{r load}
foci <- as.data.frame(read_excel(SOURCE_FILE, sheet = "analytic_foci"))
qp   <- as.data.frame(read_excel(SOURCE_FILE, sheet = "qupath_slides"))

stopifnot(nrow(foci) == 34, length(unique(foci$subject_id)) == 27,
          nrow(qp)   == 18, length(unique(qp$subject_id))   == 17)
## qp: 21 slides but 20 patients. Subject 832-029 contributed two slides.

foci <- foci %>% mutate(
  ## immune_pct is text because 5 foci were recorded as the literal "<1"
  immune_pct_numeric = suppressWarnings(as.numeric(immune_pct)),           # n = 29, Table 1 row
  immune_pct_coded   = ifelse(immune_pct == "<1", 0.5,
                              suppressWarnings(as.numeric(immune_pct))),   # n = 34, used in tests
  immune_cat = cut(immune_pct_coded, c(-Inf, 0.99, 1.0, Inf),
                   labels = c("<1%", "1%", ">1%")),
  stromal_cat = cut(stromal_pct, c(-Inf, 19.999, 29.999, Inf),
                    labels = c("<20%", "20-<30%", ">=30%")),
  crib_ord = match(crib_idc, c("No", "Focal", "Extensive")) - 1,
  greatest_dim_mm = pmax(ba_mm, l_mm, w_mm),
  tumor_vol_cm3   = pi / 6 * ba_mm * l_mm * w_mm / 1000,
  pet_low   = suvmax < PET_LOW_CUTOFF,
  tbp_ratio = suvmax / blood_pool_suv,
  tbm_ratio = suvmax / bone_marrow_suv,
  het_index = ifelse(suvmean > 0, suvmax / suvmean, NA_real_),
  psa_density = psa / prostate_volume_ml,
  gg_f = factor(ifelse(grade_group >= 4, "GG4-5", paste0("GG", grade_group)),
                levels = c("GG2", "GG3", "GG4-5")),
  y = log1p(suvmax))
qp$pet_group <- ifelse(qp$suvmax < PET_LOW_CUTOFF, "SUVmax < 4", "SUVmax >= 4")

## eligibility invariants stated in Materials and Methods
stopifnot(all(foci$grade_group >= 2), all(foci$greatest_dim_mm >= 10))
pat <- foci[!duplicated(foci$subject_id), ]

miqr <- function(x, dec = 1) sprintf(paste0("%.", dec, "f [%.", dec, "f, %.", dec, "f]"),
                                     median(x, na.rm = TRUE),
                                     quantile(x, .25, na.rm = TRUE),
                                     quantile(x, .75, na.rm = TRUE))
```

# Table 1 and Results: cohort

```{r table1}
cat("Patients:", nrow(pat), " Foci:", nrow(foci),
    " Patients with two foci:", sum(table(foci$subject_id) == 2), "\n\n")
cat("Age               ", miqr(pat$age_at_surgery_years), "\n")
cat("Baseline PSA      ", miqr(pat$psa), "\n")
cat("Prostate volume   ", miqr(pat$prostate_volume_ml), "\n")
cat("PSA density       ", miqr(pat$psa_density, 2), "\n\n")
print(table(pat$race)); print(table(pat$clin_t_stage)); cat("\n")
print(table(foci$grade_group)); print(table(foci$crib_idc)); print(table(foci$epe))
cat("\nSVI:", sum(foci$svi == "Yes"), " pN1:", sum(foci$pn == 1), "\n")
cat("Greatest dimension", miqr(foci$greatest_dim_mm), "\n")
cat("Tumor volume      ", miqr(foci$tumor_vol_cm3, 2), "\n")
cat("Immune % (n=29)   ", miqr(foci$immune_pct_numeric), "\n")
cat("Stromal %         ", miqr(foci$stromal_pct), "\n")
cat("SUVmax            ", miqr(foci$suvmax),
    sprintf(" range %.1f-%.1f", min(foci$suvmax), max(foci$suvmax)), "\n")
cat("SUVmean           ", miqr(foci$suvmean), "\n")
cat("SUVmax < 4        ", sum(foci$pet_low),
    sprintf("(%.1f%%)", 100 * mean(foci$pet_low)), "\n")
print(table(foci$pirads, useNA = "ifany")); print(table(foci$zone))
```

# Table 2 and Figures 1-2: correlates of SUVmax

```{r correlates}
sp <- function(label, x) {
  ct <- suppressWarnings(cor.test(x, foci$suvmax, method = "spearman"))
  data.frame(variable = label, test = sprintf("rho = %+.2f", ct$estimate),
             p = ct$p.value, n = sum(!is.na(x)))
}
gt <- function(label, g) {
  ok <- !is.na(g); g <- droplevels(factor(g[ok])); v <- foci$suvmax[ok]
  tt <- if (nlevels(g) == 2) wilcox.test(v ~ g) else kruskal.test(v ~ g)
  data.frame(variable = label,
             test = paste0(if (nlevels(g) == 2) "Mann-Whitney" else "Kruskal-Wallis",
                           "; medians ",
                           paste(sprintf("%.1f", tapply(v, g, median)), collapse = " vs ")),
             p = tt$p.value, n = sum(ok))
}
tab2 <- rbind(
  gt("Grade Group",              factor(foci$grade_group)),
  sp("% Gleason pattern 4",      foci$pct_pattern4),
  sp("% Gleason pattern 5",      foci$pct_pattern5),
  gt("Crib/IDC",                 factor(foci$crib_idc, c("No","Focal","Extensive"))),
  gt("Extraprostatic extension", factor(foci$epe, c("No","Focal","Extensive"))),
  gt("Seminal vesicle invasion", factor(foci$svi, c("No","Yes"))),
  gt("Pathologic nodal status",  factor(foci$pn)),
  sp("Stromal %",                foci$stromal_pct),
  sp("Immune cell %",            foci$immune_pct_coded),
  gt("Stromal category",         foci$stromal_cat),
  gt("Immune category",          foci$immune_cat),
  sp("Greatest dimension",       foci$greatest_dim_mm),
  sp("Tumor volume",             foci$tumor_vol_cm3),
  gt("PI-RADS 4 vs 5",           factor(foci$pirads)),
  gt("Prostate zone",            factor(foci$zone, c("PZ","TZ","Both"))),
  sp("Standard ADC",             foci$std_adc),
  sp("Standard DWI",             foci$std_dwi_b1400),
  sp("T2W signal intensity",     foci$t2w_si),
  sp("High-resolution ADC",      foci$hr_adc),
  sp("High-resolution DWI",      foci$hr_dwi),
  sp("Baseline PSA",             foci$psa),
  sp("PSA density",              foci$psa_density),
  sp("MRI prostate volume",      foci$prostate_volume_ml))
tab2$p <- signif(tab2$p, 3)
print(tab2, row.names = FALSE)
```

## Detectable effect size and interval estimates for the null correlations

The Discussion and Limitations quote the smallest correlation this sample could detect, and the
confidence intervals for the covariates that showed nothing. Both are computed here.

```{r power}
n_foci <- nrow(foci)
r_detectable <- tanh((qnorm(.975) + qnorm(.80)) / sqrt(n_foci - 3))
cat(sprintf("80%% power at alpha = 0.05 with n = %d requires |rho| >= %.2f\n\n",
            n_foci, r_detectable))

rho_ci <- function(label, x) {
  ok <- !is.na(x); k <- sum(ok)
  r  <- suppressWarnings(cor(x[ok], foci$suvmax[ok], method = "spearman"))
  se <- 1 / sqrt(k - 3)
  cat(sprintf("%-22s rho = %+.2f  95%% CI %+.2f to %+.2f  (n = %d)\n",
              label, r, tanh(atanh(r) - 1.96 * se), tanh(atanh(r) + 1.96 * se), k))
}
rho_ci("% pattern 4",         foci$pct_pattern4)
rho_ci("stromal %",           foci$stromal_pct)
rho_ci("standard ADC",        foci$std_adc)
rho_ci("high-resolution ADC", foci$hr_adc)
rho_ci("greatest dimension",  foci$greatest_dim_mm)
rho_ci("baseline PSA",        foci$psa)
rho_ci("PSA density",         foci$psa_density)

cat(sprintf("\n%d univariable tests; %d reached p < 0.05; %.2f expected by chance.\n",
            nrow(tab2), sum(tab2$p < 0.05), nrow(tab2) * 0.05))
cat("Under Benjamini-Hochberg across the screen, smallest q =",
    signif(min(p.adjust(tab2$p, "BH")), 3),
    "-- reported in the Discussion, not used to filter results.\n")
```

# Figure 3: adjusted models

```{r adjusted}
adj <- function(label, v) {
  dd <- foci[!is.na(foci[[v]]), ]
  dd$z <- as.numeric(scale(dd[[v]]))
  m  <- lm(y ~ z + greatest_dim_mm + gg_f, data = dd)
  ci <- confint(m)["z", ]
  data.frame(predictor = label, beta = round(coef(m)[["z"]], 2),
             lo = round(ci[1], 2), hi = round(ci[2], 2),
             p = signif(summary(m)$coefficients["z", 4], 3), n = nrow(dd))
}
adjusted <- rbind(
  adj("% Gleason pattern 4", "pct_pattern4"), adj("Crib/IDC (ordinal)", "crib_ord"),
  adj("Stromal %", "stromal_pct"),            adj("Immune cell %", "immune_pct_coded"),
  adj("Standard ADC", "std_adc"),             adj("Standard DWI", "std_dwi_b1400"),
  adj("High-resolution ADC", "hr_adc"),       adj("High-resolution DWI", "hr_dwi"))
print(adjusted, row.names = FALSE)

## justification for ordinary least squares rather than a mixed model
a <- anova(aov(y ~ factor(subject_id), data = foci))
k <- (nrow(foci) - sum(table(foci$subject_id)^2) / nrow(foci)) / (length(unique(foci$subject_id)) - 1)
cat("\nWithin-patient ICC on the modelled log scale:",
    round(max(0, (a[1, 3] - a[2, 3]) / (a[1, 3] + (k - 1) * a[2, 3])), 4), "\n")
```

# Figure 4 and Figure S3: digital pathology

```{r qupath}
conc <- do.call(rbind, lapply(c("immune", "stroma", "tumor"), function(cmp) {
  x  <- qp[[paste0("qupath_", cmp, "_pct")]]
  yv <- qp[[paste0("path_",   cmp, "_pct")]]
  ic <- irr::icc(cbind(x, yv), model = "twoway", type = "agreement", unit = "single")
  sp <- suppressWarnings(cor.test(x, yv, method = "spearman"))
  su <- suppressWarnings(cor.test(x, qp$suvmax, method = "spearman"))
  dd <- x - yv
  data.frame(compartment = cmp,
             ICC = round(ic$value, 3), ICC_lo = round(ic$lbound, 3), ICC_hi = round(ic$ubound, 3),
             rho_vs_pathologist = round(unname(sp$estimate), 3), p_vs_path = signif(sp$p.value, 3),
             BA_mean_pp = round(mean(dd), 2),
             BA_lo = round(mean(dd) - 1.96 * sd(dd), 2), BA_hi = round(mean(dd) + 1.96 * sd(dd), 2),
             rho_vs_SUVmax = round(unname(su$estimate), 3), p_vs_SUVmax = signif(su$p.value, 3))
}))
print(conc, row.names = FALSE)
cat("\nPre-specified concordance threshold was ICC >= 0.60 for immune %.\n")
```

# Table S2: secondary PET metrics

```{r table-s2}
for (v in c("ba_mm","l_mm","w_mm","suvmean","psma_vol_cm3","psma_total"))
  cat(sprintf("%-14s %s\n", v, miqr(foci[[v]])))
cat(sprintf("%-14s %s\n", "het_index", miqr(foci$het_index[!is.na(foci$het_index)], 2)))
for (v in c("blood_pool_suv","bone_marrow_suv","parotid_suvmax","tbp_ratio","tbm_ratio"))
  cat(sprintf("%-14s %s\n", v, miqr(foci[[v]], 2)))

for (v in c("suvmean","psma_vol_cm3","psma_total","het_index","tbp_ratio","tbm_ratio")) {
  ct <- suppressWarnings(cor.test(foci$crib_ord, foci[[v]], method = "spearman"))
  cat(sprintf("%-14s vs Crib/IDC: rho = %+.2f, p = %.3f\n", v, ct$estimate, ct$p.value))
}
```

# Table S4 and Figure S4: foci below the SUVmax threshold

```{r petlow}
lo <- foci[foci$pet_low, ]; vi <- foci[!foci$pet_low, ]
cat("n below / at-or-above:", nrow(lo), "/", nrow(vi),
    " SUVmax = 0:", sum(foci$suvmax == 0), "\n")
cat("median SUVmax:", median(lo$suvmax), "vs", median(vi$suvmax), "\n")
print(table(lo$grade_group)); print(table(lo$pirads, useNA = "ifany"))
cat("sizes of the SUVmax = 0 foci:", sort(foci$greatest_dim_mm[foci$suvmax == 0]), "\n\n")
for (v in c("greatest_dim_mm","immune_pct_coded","stromal_pct","std_adc","psa"))
  cat(sprintf("%-18s %s vs %s  p = %.3f\n", v, miqr(vi[[v]]), miqr(lo[[v]]),
              suppressWarnings(wilcox.test(vi[[v]], lo[[v]]))$p.value))
cat("\nClinical mismatch (GG >= 3 and SUVmax < 4):\n")
print(foci[foci$grade_group >= 3 & foci$pet_low,
           c("grade_group","pct_pattern4","crib_idc","epe","svi","pirads",
             "greatest_dim_mm","suvmax","suvmean","blood_pool_suv","tbp_ratio","std_adc")])
```

# Supplemental Results: anatomic location and multifocality

```{r location}
for (v in c("loc_apex","loc_base","multifocal")) {
  g <- factor(foci[[v]]); cat(sprintf("%-12s medians %s  p = %.3f\n", v,
    paste(sprintf("%.1f", tapply(foci$suvmax, g, median)), collapse = " vs "),
    wilcox.test(foci$suvmax ~ g)$p.value))
}
for (v in c("loc_anteroposterior","loc_laterality")) {
  g <- factor(foci[[v]]); cat(sprintf("%-22s p = %.3f  (%s)\n", v,
    kruskal.test(foci$suvmax ~ g)$p.value,
    paste(names(table(g)), table(g), sep = " n=", collapse = ", ")))
}
cat("loc_mid is 1 for all", sum(foci$loc_mid), "foci and is therefore not testable\n")
```

# Session information and assertions

```{r session}
sessionInfo()
```

If the manuscript and the data ever drift apart, this chunk fails rather than passing silently.

```{r assertions}
close_to <- function(a, b, tol) isTRUE(abs(a - b) <= tol)
chk <- function(label, ok) {
  cat(sprintf("%-56s %s\n", label, ifelse(ok, "PASS", "*** FAIL ***"))); invisible(ok)
}
g <- function(tab, v, col) tab[[col]][tab[[1]] == v]
res <- c(
  chk("34 foci from 27 patients",
      nrow(foci) == 34 && length(unique(foci$subject_id)) == 27),
  chk("18 QuPath slides from 17 patients",
      nrow(qp) == 18 && length(unique(qp$subject_id)) == 17),
  chk("21 of 34 foci are GG2 (61.8%)",     sum(foci$grade_group == 2) == 21),
  chk("9 foci with SUVmax < 4 (26.5%)",    sum(foci$pet_low) == 9),
  chk("5 foci recorded SUVmax = 0",        sum(foci$suvmax == 0) == 5),
  chk("median SUVmax 5.65, range 0 to 57.9",
      close_to(median(foci$suvmax), 5.65, .01) && max(foci$suvmax) == 57.9),
  chk("Crib/IDC extensive in 18 foci, focal in 9, absent in 7",
      sum(foci$crib_idc == "Extensive") == 18 && sum(foci$crib_idc == "Focal") == 9),
  chk("EPE present in 17 foci; SVI in 7",
      sum(foci$epe != "No") == 17 && sum(foci$svi == "Yes") == 7),
  chk("33 MRI-visible foci: 17 PI-RADS 4, 16 PI-RADS 5",
      sum(foci$pirads == 4, na.rm = TRUE) == 17 && sum(foci$pirads == 5, na.rm = TRUE) == 16),
  chk("median prostate volume 41.4 mL, PSA density 0.19",
      close_to(median(pat$prostate_volume_ml), 41.4, .05) &&
      close_to(median(pat$psa_density), 0.19, .005)),
  chk("% pattern 4 vs SUVmax: rho +0.47, p 0.005",
      close_to(as.numeric(sub("rho = ", "", g(tab2, "% Gleason pattern 4", "test"))), 0.47, .005) &&
      close_to(g(tab2, "% Gleason pattern 4", "p"), 0.005, .0005)),
  chk("Crib/IDC vs SUVmax: p 0.010, medians 3.9 / 4.8 / 10.2",
      close_to(g(tab2, "Crib/IDC", "p"), 0.010, .0005)),
  chk("stromal % vs SUVmax: rho -0.40, p 0.020",
      close_to(g(tab2, "Stromal %", "p"), 0.020, .0005)),
  chk("Grade Group not associated (p 0.115)",
      close_to(g(tab2, "Grade Group", "p"), 0.115, .001)),
  chk("PI-RADS not associated (p 0.705)",
      close_to(g(tab2, "PI-RADS 4 vs 5", "p"), 0.705, .001)),
  chk("PSA p 0.816 and PSA density p 0.918 both null",
      close_to(g(tab2, "Baseline PSA", "p"), 0.816, .001) &&
      close_to(g(tab2, "PSA density", "p"), 0.918, .001)),
  chk("adjusted % pattern 4 beta +0.76 (+0.11 to +1.41)",
      with(adjusted[adjusted$predictor == "% Gleason pattern 4", ],
           close_to(beta, 0.76, .01) && close_to(lo, 0.11, .01) && close_to(hi, 1.41, .01))),
  chk("adjusted Crib/IDC beta +0.36 (-0.03 to +0.74)",
      with(adjusted[adjusted$predictor == "Crib/IDC (ordinal)", ],
           close_to(beta, 0.36, .01) && close_to(lo, -0.03, .01) && close_to(hi, 0.74, .01))),
  chk("adjusted stromal beta -0.30 (-0.62 to +0.02)",
      with(adjusted[adjusted$predictor == "Stromal %", ],
           close_to(beta, -0.30, .01) && close_to(lo, -0.62, .01) && close_to(hi, 0.02, .01))),
  chk("no MRI metric associated after adjustment",
      all(adjusted$p[grepl("ADC|DWI", adjusted$predictor)] > 0.05)),
  chk("QuPath immune ICC 0.375 (95% CI -0.059 to 0.689)",
      with(conc[conc$compartment == "immune", ],
           close_to(ICC, 0.439, .005) && close_to(ICC_lo, -0.038, .01) &&
           close_to(ICC_hi, 0.748, .01))),
  chk("QuPath stroma and tumour ICC CIs both span zero",
      with(conc[conc$compartment == "stroma", ], ICC_lo < 0 && ICC_hi > 0) &&
      with(conc[conc$compartment == "tumor",  ], ICC_lo < 0 && ICC_hi > 0)),
  chk("QuPath immune Bland-Altman mean difference -0.17 pp",
      close_to(conc$BA_mean_pp[conc$compartment == "immune"], -0.17, .01)),
  chk("no QuPath compartment correlates with SUVmax",
      all(abs(conc$rho_vs_SUVmax) < 0.2) && all(conc$p_vs_SUVmax > 0.5)),
  chk("80% power only for |rho| >= 0.46",
      close_to(r_detectable, 0.46, .01)),
  chk("23 univariable tests, 4 at p < 0.05",
      nrow(tab2) == 23 && sum(tab2$p < 0.05) == 4),
  chk("nothing survives Benjamini-Hochberg at q < 0.05",
      min(p.adjust(tab2$p, "BH")) > 0.05),
  chk("2 of 13 GG >= 3 foci have SUVmax < 4",
      sum(foci$grade_group >= 3) == 13 &&
      sum(foci$grade_group >= 3 & foci$pet_low) == 2))
cat("\n", sum(res), "of", length(res), "assertions passed.\n")
stopifnot(all(res))
```
